Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-07-29 paper/main.tex
relabel

%###################################################################################################
\section{Methodology}
% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation of Rod-Shaped Bacteria}
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{x}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{x}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{x}_{i-1},\vec{x}_i$ can be used when modeling the
dynamics of these bacteria.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Visualization}
\label{section:generating-masks-and-images}

@@

In order to determine if our extraction algorithm provides reliable results, we have to take
advantage of a functionality which will be introduced later in
section~\ref{section:generating-masks-and-images}.
Our methodology makes use of the fact that we have already set up our model and solving scheme and
are able to produce cell masks from these numerical results.
In doing so, we have exact knowledge about the position of our agents since we can simply obtain
this information from our model.
But on the other hand we can also try to extract this position from the generated cell-masks.
By comparing the differences between them, we get a good estimate of how well our extraction
algorithm can determine the true position of the bacteria.

>>>>

%###################################################################################################
\section{Methodology}
% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation of Rod-Shaped Bacteria}
\label{subsec:spatial-representation}
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{x}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{x}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{x}_{i-1},\vec{x}_i$ can be used when modeling the
dynamics of these bacteria.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Visualization}
\label{subsubsection:visualization}

@@

In order to determine if our extraction algorithm provides reliable results, we have to take
advantage of a functionality which will be introduced later in
section~\ref{subsubsection:visualization}.
Our methodology makes use of the fact that we have already set up our model and solving scheme and
are able to produce cell masks from these numerical results.
In doing so, we have exact knowledge about the position of our agents since we can simply obtain
this information from our model.
But on the other hand we can also try to extract this position from the generated cell-masks.
By comparing the differences between them, we get a good estimate of how well our extraction
algorithm can determine the true position of the bacteria.
